A NetdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm

### 2024-12-06 paper.qmd
Adding final edits

Where $W_{ij}$ is a weight matrix indicating ties between individuals, most commonly a binary adjacency matrix such as who is friends with whom but can be any relation; a is an adoption vector indicating which of i’s j alters had adopted at time t-1; and the denominator is the row sum, the number of outgoing ties. Eit in (1) measures the proportion of a person’s ties at time t who have adopted the behavior at the prior time period, which is a contagion measure. As described in @Valente2024, however, there can be many extensions of the exposure calculation by estimating influence of ties based on: a count (removing the denominator), indirect connectivity, structural equivalence, strength (weak vs. strong), homophily, Simmelian relations (alter-to-alter), alter network properties such as centrality, etc. NetdiffuseR enables users to calculate these alternative exposures with a parameter toggle in the exposure function.

@@

As you can see, the dataset contains 1,047 observations with 432 columns. The dataset consists of survey data collected from women of childbearing age in 25 rural villages. For this example, we will focus on the two largest villages, namely villages 2 and 21. We can subset the data using the following code:

@@

The survey has multiple network nomination questions. Depending on the situation, you may want to aggregate them or focus on one of them. We will use the “talks to a neighbor” network nomination questions, where participants nominated up to five people. Using that information, we can now use the `survey_to_diffnet` function to create a diffnet object:

@@

The key things to notice from this code are: (1) We have to specify the “id” of each individual in the dataset via the `idvar` variable; (2) using the `netvars` argument, you can list all the network nomination columns in your data; (3) the `toavar` column is the name of the variable that contains the time of adoption (year in this case); finally (4) since we have multiple communities in this dataset, we use the `groupvar` parameter to tell the function there cannot be ties between villages, only within. The function has many more options and readers can learn about them in the manual `?survey_to_diffnet`.

@@

```{r}
#| label: diffreg-2
#| warning: false
diffreg(
  kfam_diffnet ~ age + factor(village) +
    exposure(
      alt.graph = exp_age_homo_graph,
      valued = TRUE
      )
  ) |>
  summary()
```

As you can see, we can pass arguments to the `exposure` function on the fly, including using the R's formula interface which is commonly used with the `glm` function in the package. More details about what happens when calling the `diffreg` function can be found in the function's manual.

@@

, which is equivalent to calling `rdiffnet(200, 10)`:

@@

Starting version 1.24.0, **netdiffuseR** supports simulating multiadoption diffusion and disadoption processes. To select a multiadoption process, you can pass a `list` as the `seed.p.adopt` parameter. Here is a simple example using the Koren Family Planning network:

```{r}
#| label: simulation-multiadoption-process-4
diffnet_4 <- rdiffnet(
  200, 10,
  seed.p.adopt = list(0.1, 0.15)
  )

@@

Inspecting the output from the print method of the `diffnet` object, we can see that the object contains two behaviors: The  "Num of behaviors" entry now shows `2`, the "Behavior" entry also shows two behaviors, `"Random contagion \_1, Random contagion\_2"`, and finally, the "Prevalence" entry also shows two numbers: `0.92, 0.96`. Although we are simulating two behaviors, `rdiffnet` will simulate as many as values are in the `seed.p.adopt` list.

::: {.callout-tip}
In the current implementation of `rdiffnet`, the multiadoption module simulates behaviors independently. That is, the code above would be equivalent to simulating the same behavior twice. More complicated models in which behaviors are interdependent are supported via the disadoption parameter.
:::

@@

In this particular example, we ran `rdiffnet` with most of the parameters being `list`. The reader is invited to look at other types of possible inputs in the `rdiffnet` documentation.

@@

An template for a disadoption function, which currently returns no disadoption, follows:

```{r}
#| label: template-disadoption
#| eval: false
disadoption_function <- function(expo, cumadopt, time) {

  list(integer(), integer())

@@

  # Number of behaviors
  num_of_behaviors <- dim(cumadopt)[3]

  # Making room for the disadopted nodes
  list_disadopt <- list(integer(), integer())

  # We iterate through the behaviors
  for (q in 1:num_of_behaviors) {

@@

  return(list_disadopt)
}
```

@@

3. The `time` argument is the current time point in the simulation. This allows the function to know when to disadopt function is beign called.

@@

This paper has provided an introduction to the NetdiffuseR R package designed to facilitate estimating contagion and diffusion processes on networks. The main goal was to provide example code and tutorial so others may use the package to explore diffusion network models either with the existing classic datasets that come with the package or using their own data. The main tutorial sections included sample code for estimating network exposure, the workhorse variable in network research. We also provided examples of extensions to exposure that provide a wide array of different weighting techniques, such as structural equivalence, homophilic ties, and so on.

>>>>

Where $W_{ij}$ is a weight matrix indicating ties between individuals, most commonly a binary adjacency matrix such as who is friends with whom but can be any relation; $a$ is an adoption vector indicating which of $i$’s $j$ alters had adopted at time $t - 1$; and the denominator is the row sum, the number of outgoing ties. $E_{it}$ in (1) measures the proportion of a person’s ties at time $t$ who have adopted the behavior at the prior time period, which is a contagion measure. As described in @Valente2024, however, there can be many extensions of the exposure calculation by estimating influence of ties based on: a count (removing the denominator), indirect connectivity, structural equivalence, strength (weak vs. strong), homophily, Simmelian relations (alter-to-alter), alter network properties such as centrality, etc. NetdiffuseR enables users to calculate these alternative exposures with a parameter toggle in the exposure function.

@@

As you can see, the dataset contains 1,047 observations with 432 columns. The dataset consists of survey data collected in 1973-1974 from women of childbearing age in 25 rural villages. For this example, we will focus on the two largest villages, namely villages 2 and 21. We can subset the data using the following code:

@@

The survey has multiple network nomination questions. Depending on the situation, you may want to aggregate them or focus on one of them. We will use the “talks to a neighbor” network nomination question, where participants nominated up to five people. Using that information, we can now use the `survey_to_diffnet` function to create a diffnet object:

@@

The key things to notice from this code are: (1) We have to specify the “id” of each individual in the dataset via the `idvar` argument; (2) using the `netvars` argument, you can list all the network nomination columns in your data; (3) the `toavar` argument is the name of the variable that contains the time of adoption (year in this case); finally (4) since we have multiple communities in this dataset, we use the `groupvar` argument to tell the function there cannot be ties between villages, only within. The function has many more options and readers can learn about them in the manual `?survey_to_diffnet`.

@@

```{r}
#| label: diffreg-2
#| warning: false
diffreg(
  kfam_diffnet ~ age + factor(village) +
    exposure(
        lag = 1,
        alt.graph = exp_age_homo_graph,
        valued = TRUE
        )
  ) |>
  summary()
```

As you can see, we can pass arguments to the `exposure` function on the fly, including using the R's formula interface which is commonly used with the `glm` function in the package. More details about what happens when calling the `diffreg` function can be found in the library's manual.

@@

 which is equivalent to calling `rdiffnet(200, 10)`:

@@

Starting with version 1.24.0, **netdiffuseR** supports simulating multi- and dis-adoption diffusion processes. To study a multi-adoption process, you can pass a `list` as the `seed.p.adopt` parameter. Here is a simple example using the Koren Family Planning network:

```{r}
#| label: simulation-multiadoption-process-4
set.seed(1231)

diffnet_4 <- rdiffnet(
 200, 10,
 seed.p.adopt = list(0.1, 0.15)
 )

@@

Inspecting the output from the print method of the `diffnet` object, we can see that the object contains two behaviors: The  "Num of behaviors" entry now shows `2`, the "Behavior" entry also shows two behaviors, `"Random contagion \_1, Random contagion\_2"`, and finally, the "Prevalence" entry also shows two numbers: `0.29, 0.97`. Although we are simulating two behaviors, `rdiffnet` will simulate as many as values are in the `seed.p.adopt` list.

::: {.callout-tip}
In the current implementation of `rdiffnet`, the multi-adoption module simulates behaviors independently. That is, the code above would be equivalent to simulating the same behavior twice. More complicated models in which behaviors are interdependent are supported via the dis-adoption parameter.
:::

@@

In this particular example, we ran `rdiffnet` with most of the parameters being in a `list`. The reader is invited to look at other types of possible inputs in the `rdiffnet` documentation.

@@

A template for a disadoption function, which currently returns no disadoption, follows:

```{r}
#| label: template-disadoption
#| eval: false
disadoption_function <- function(expo, cumadopt, time) {
  list(integer(), integer())
}
```

@@

  # Number of behaviors
  num_of_behaviors <- dim(cumadopt)[3]
    # Making room for the disadopted nodes
  list_disadopt <- list(integer(), integer())
    # We iterate through the behaviors
  for (q in 1:num_of_behaviors) {

@@

 return(list_disadopt)

@@

3. The `time` argument is the current time point in the simulation. This allows the function to know when the disadopt function is being called.

@@

This paper has provided an introduction to the NetdiffuseR R package designed to facilitate estimating contagion and diffusion processes on networks. The main goal was to provide example code and tutorial so others may use the package to explore diffusion network models either with the existing classic datasets that come with the package or using their own data. The main tutorial sections included sample code for estimating network exposure, the workhorse variable in network research. Simulation features of netdiffuseR were illustrated which provide tools for theoretical testing of diffusion processes. We also provided examples of extensions to exposure that provide a wide array of different weighting techniques, such as structural equivalence, homophilic ties, and so on.
